Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-12-05 draft.tex
shortened 'Experiments and Preliminary Results' section

\noindent Spine shape is possibly associated with several somatometric variables.
%% TODO: Shorten

@@

Since men are on average taller than women and people of old age shrink due to \emph{bone erosion}, \emph{gender} and \emph{age} are also risk factors.

@@

Women have a higher life expectancy than men, hence their higher share in the old age group.

@@

Also, women are on average smaller than men, hence the larger shape similarity with older subjects.

@@

Large \emph{body weight} increases the spine load, resulting in a bent shape.

@@

As a first experiment we correlated the shape variable with the dichotomous back pain indicator using \texttt{GPLOM's} \cite{GPLOMS}.

@@

Since our image-derived variables are metric, their pairwise combinations are visualized using scatter plots.

@@

@@

We conclude that \emph{curvature} is not as reliable on the transversal plane as it is on the other planes.

@@

\noindent \textbf{Correlation Matrix.}

@@

We calculated a correlation matrix to assess correlations between the image variables.

@@

The pearson correlation coefficients between the numeric variables are depicted right of the matrix diagonal (Fig.~\ref{fig:image-parameter-range}).

@@

\emph{Curvature}, \emph{curvature angle} and \emph{torsion} strongly correlate with their planar projections.
Also the \emph{mean curvature} and \emph{curvature angle} correlate by a factor of $-0.89$.
\emph{Torsion} does not correlate with any other image-derived variable.
\noindent \textbf{Correlation of Image Variables With Lumbar Back Pain.}
Figure~\ref{fig:image-parameter-range} shows the distribution of all image-variables as Generalized Pairs Plot.
No statistically significant correlation could be observed through all subject groups.
The box plots show no difference between subjects with or without back pain.
\noindent \textbf{Assessing the Information Gain Using the Principal Component Analysis.}
To determine the information gain per image-derived variable, we calculated a Principal Component Analysis (PCA) and compared the dimension loadings.
The PCA results in a linear projection of the variable space ordering their components by maximum variance.
The first three principal components explain 75\% of the variance in the nine image-derived variables.
The first principal component explains 47\% of the variance and comprises primarily \emph{mean curvature}, \emph{sagittal curvature}, \emph{curvature angle} and \emph{sagittal curvature angle}.
The second component, adding 16\% of the variance, weigths \emph{coronal curvature} and \emph{coronal curvature angle}.
The third component explains 12\% of the variance and comprises \emph{torsion} and \emph{transverse curvature}.
This supports our prior conclusion about the low information gain of the transverse planes.
\emph{Torsion} also adds little variance to the information space.
The complete analysis can be found in the experiments section of the supplementary material (\url{blind.dnsalias.com})

@@

The method uses the following correlation metrics for the different type combinations:

@@

@@

\item \emph{Pearson product-moment} for two continuous variables, 
\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
\item \emph{Polychoric} correlations for two categorical variables.

@@

All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - perfect correlation}.

@@

Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
These variables have to be omitted in the analysis, since they are not statistically resilient.
We display the resulting \emph{contingency matrix} using a heat map, encoding correlation values with color brightness with white for $0$ and dark blue for $1$, as presented in \cite{Klemm2014VIS}.

@@

\vfill
\bibliographystyle{style/apalike}
{\small
\bibliography{bibliography}}

>>>>

\noindent Spine shape is influenced by several somatometric variables.
%% TODO: Introduce this better! Go into detail on the train of thought

@@

%Since men are on average taller than women and people of old age shrink due to \emph{bone erosion}, \emph{gender} and \emph{age} are also risk factors.

@@

%Women have a higher life expectancy than men, hence their higher share in the old age group.

@@

%Also, women are on average smaller than men, hence the larger shape similarity with older subjects.

@@

High \emph{body weight} increases the spine load, resulting in a bent shape.

@@

As a first experiment we correlated the shape variable with the dichotomous back pain indicator using a \texttt{GPLOM}.

@@

The metric image-derived variables are pairwise visualized using scatter plots.
%Since our image-derived variables are metric, their pairwise combinations are visualized using scatter plots.

@@

%% % TODO: [high] Add Link
We conclude that \emph{curvature} is not as reliable on the transversal plane as it is on the other planes, which was also confirmed through a Principal Component Analysis (see supplementary material at TODO:LINK).

@@

The \texttt{GPLOM} shows very similar distributions of subjects with or without back pain with respect to the shape variables in all sub plots.

@@

% \\\\
% \noindent \textbf{Correlation Matrix.}

@@

% We calculated a correlation matrix to assess correlations between the image variables.
% The pearson correlation coefficients between the numeric variables are depicted right of the matrix diagonal (Fig.~\ref{fig:image-parameter-range}).
% \emph{Curvature}, \emph{curvature angle} and \emph{torsion} strongly correlate with their planar projections.
% Also the \emph{mean curvature} and \emph{curvature angle} correlate by a factor of $-0.89$.
% \emph{Torsion} does not correlate with any other image-derived variable.
% \\\\
% \noindent \textbf{Correlation of Image Variables With Lumbar Back Pain.}
% Figure~\ref{fig:image-parameter-range} shows the distribution of all image-variables as Generalized Pairs Plot.
% No statistically significant correlation could be observed through all subject groups.
% The box plots show no difference between subjects with or without back pain.
% \\\\
% \noindent \textbf{Assessing the Information Gain Using the Principal Component Analysis.}
% A Principal Component Analysis (PCA) of the image data yielded that \emph{mean curvature}, \emph{sagittal curvature}, \emph{curvature angle} and \emph{sagittal curvature angle} explain almost 50\% of the image variance.
% \emph{Torsion} and metrics of the \emph{transverse} projection planes contribute very little variance, supporting our conclusion about their low information gain.

@@

@@

% To determine the information gain per image-derived variable, we calculated a Principal Component Analysis (PCA) and compared the dimension loadings.
% The PCA results in a linear projection of the variable space ordering their components by maximum variance.
% The first three principal components explain 75\% of the variance in the nine image-derived variables.
% The first principal component explains 47\% of the variance and comprises primarily \emph{mean curvature}, \emph{sagittal curvature}, \emph{curvature angle} and \emph{sagittal curvature angle}.
% The second component, adding 16\% of the variance, weigths \emph{coronal curvature} and \emph{coronal curvature angle}.
% The third component explains 12\% of the variance and comprises \emph{torsion} and \emph{transverse curvature}.
% This supports our prior conclusion about the low information gain of the transverse planes.
% \emph{Torsion} also adds little variance to the information space.
% The complete analysis can be found in the experiments section of the supplementary material (\url{blind.dnsalias.com})

@@

The method uses \emph{pearson product-moment} for two continuous variables, \emph{polyserial} correlation for one continuous and one categorical variable, and \emph{Polychoric} correlations for two categorical variables.
%The method uses the following correlation metrics for the different type combinations:

@@

% \begin{itemize}
% \item \emph{Pearson product-moment} for two continuous variables,
% \item \emph{Polyserial} correlation for one continuous and one categorical variable, and
% \item \emph{Polychoric} correlations for two categorical variables.
% \end{itemize}

@@

All values are scaled between \texttt{0 - no correlation} and \texttt{1 - perfect correlation}.

@@

Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure} and are omitted, since they are not statistically resilient.

@@

We display the resulting \emph{contingency matrix} using a heat map, encoding correlation values with color brightness with white for $0$ and dark blue for $1$.

@@

% \vfill
\bibliographystyle{style/apalike}
{\small
\bibliography{bibliography}}
